---
title: "Darwin Core mapping"
subtitle: "For: Ad hoc checklist of alien species in Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here]().

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse)       # To transform data
library(magrittr)        # For %<>% pipes
library(googlesheets)    # To import and read Google spreadsheets 
library(stringr)         # To perform string operations
library(digest)          # To generate hashes
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/ad_hoc_checklist_dump.csv"

# Processed files:

```

# Read the raw data using googlesheets

The original spreadsheet can be found [here](https://docs.google.com/spreadsheets/d/1LeXXbry2ArK2rngsmFjz_xErwE1KwQ8ujtvHNmTVA6E/edit#gid=0). We need to retrieve this spreadsheet and select the specific worksheet first:

Retrieve the spreadsheet:

```{r}
retrieve_spreadsheet <- gs_title("ad hoc checklist")
```

select the data in the worksheet `checklist`:

```{r}
raw_data <- retrieve_spreadsheet %>% gs_read("checklist")
```

We want to add a copy of the ad hoc checklist to the Github repository:

```{r}
write.csv(raw_data, file = raw_data_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```
